---
title: "Analysis of How Do Temperature, Sugar Type, and Stirring Affect Sugar Dissolution Time"
author: "Sakura Hu"
output:
  pdf_document:
    number_sections: true
    fig_caption: true
  html_document:
    number_sections: true
    fig_caption: true
thanks: "Code and data are available at: [https://github.com/xycww/sta305_sugar](https://github.com/xycww/sta305_sugar)."
---

# Description of the design
This experiment investigates how three factors—water temperature, sugar type, and stirring—affect the dissolution time of granulated sugar. A replicated $2^3$ full factorial design was used, involving three factors at two levels each, resulting in eight treatment combinations. Each combination was replicated three times, yielding a total of 24 trials. This design was chosen because it offers a comprehensive yet manageable framework to examine both main effects and interaction effects, while enabling replication to estimate experimental variability.

The selected factors represent distinct mechanisms influencing solubility. 

- **Water Temperature**: cold tap water vs. hot water at 70°C. This variable is chosen since it affects molecular motion and reaction rates (thermal effects). 

- **Sugar Type**: White granulated sugar vs. brown granulated sugar. This variable is chosen since it influences dissolution via chemical composition and crystal structure.

- **Stirring**: No stirring vs. stirring at a consistent speed (approximately two full rotations per second using a spoon). This variable is chosen since it could accelerate solute dispersion.

To conduct the experiment, 450 mL of water was measured into a clear container for each trial, with temperature conditions prepared using either cold tap water or freshly boiled water cooled briefly. One teaspoon of the designated sugar type was added, and a stopwatch was started simultaneously. For stirring conditions, the solution was stirred by hand in a consistent circular motion; otherwise, no external movement was applied. Timing stopped once all visible granules were dissolved, and the result was recorded in seconds. Trials were performed in randomized order to mitigate systematic bias.

This setup was selected due to its feasibility under individual constraints: it requires no specialized equipment or human participants, yet still allows for full experimental control. Each factor is clearly defined, discretely manipulable, and easily repeated, and the response variable—**time to full dissolution**—is straightforward to quantify.

Beyond practicality, the experiment serves as a model for broader physical processes, particularly in contexts such as food preparation, beverage formulation, and industrial solubility studies. The factorial design allows not only for the assessment of individual effects (e.g., temperature alone) but also complex interactions—such as whether the effect of stirring depends on temperature, or whether brown sugar dissolves disproportionately slower in cold water without agitation. Identifying such interactions is often more informative than studying single factors in isolation, and the results of this design may provide insight into how simple environmental variables combine to influence a commonly encountered chemical process.

# Analysis of the data
```{r, include=FALSE}
#load all libraries
library(tidyverse)
library(FrF2)
library(dplyr)
library(ggplot2)
```

```{r, include=FALSE}
#load the data
data <- read.csv("../data/sugar.csv")
head(data)
```

A full factorial $2^3$ design was implemented with three replicates per treatment combination, yielding 24 observations. The response variable, dissolution time (in seconds), was measured across combinations of three binary factors: Temperature (Cold vs. Hot), Sugar Type (White vs. Brown), and Stirring (No vs. Yes). Factor levels were coded as follows to simplify the construction and interpretation of the linear model:

\begin{itemize}
  \item \textbf{Temperature}: $-1$ for cold water, $+1$ for hot water.
  \item \textbf{Sugar Type}: $-1$ for brown sugar, $+1$ for white sugar.
  \item \textbf{Stirring}: $-1$ for no stirring, $+1$ for stirring.
\end{itemize}

The factorial structure of the experiment enables estimation of all main effects, two-way interactions, and the three-way interaction, as well as an estimate of experimental error.

## Model
The response was modeled using the standard linear model for a $2^3$ factorial design:

\[
y_i = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_3 + \beta_4 x_1 x_2 + \beta_5 x_1 x_3 + \beta_6 x_2 x_3 + \beta_7 x_1 x_2 x_3 + \varepsilon_i,
\]

where:
\begin{itemize}
  \item $x_1$, $x_2$, and $x_3$ represent coded levels of Temperature, Sugar Type, and Stirring, respectively,
  \item $\beta_0$ is the overall mean response,
  \item $\beta_1$, $\beta_2$, $\beta_3$ are the main effects of Temperature, Sugar Type, and Stirring,
  \item $\beta_4$, $\beta_5$, $\beta_6$ represent two-way interactions (e.g., Temperature $\times$ Sugar),
  \item $\beta_7$ represents the three-way interaction between all factors,
  \item $\varepsilon_i \sim \mathcal{N}(0, \sigma^2)$ represents random errors assumed to follow a normal distribution with mean 0 and variance $\sigma^2$.
\end{itemize}

```{r, include=FALSE}
# Recode factors into numeric coded levels (-1, +1)
data$x1 <- ifelse(data$Temp == "hot", 1, -1)
data$x2 <- ifelse(data$Sugar == "white", 1, -1)
data$x3 <- ifelse(data$Stirring == "yes", 1, -1)

# Interaction terms
data$x4 <- data$x1 * data$x2
data$x5 <- data$x1 * data$x3
data$x6 <- data$x2 * data$x3
data$x7 <- data$x1 * data$x2 * data$x3

# Fitting model
model_coded <- lm(Time_to_Dissolve ~ x1 + x2 + x3 + x4 + x5 + x6 + x7, data = data)
summary(model_coded)

```

## Main Effects, Interactions, and Significance

The model was fitted to obtain the estimated coefficients and corresponding p-values. The factorial effects for each factor and interaction are estimated by calculating $2\hat{\beta}_j$, reflecting the difference between high and low levels of each factor.

Given $m = 3$ replicates and $k = 3$ factors, and with the pooled estimate of residual variance $\hat{\sigma}^2 = 65.12^2 = 4240.06$, the variance of each factorial effect was computed as:
\[
\text{Var(effect)} = \frac{4 \hat{\sigma}^2}{m \cdot 2^k} = \text{706.77}
\]
The 95%confidence interval of each effect is calculated as 
\[
2\hat{\beta}_j \pm t_{0.975, 16} \cdot \sqrt{\text{Var(effect)}}.
\]
An effect was considered statistically significant if its p-value was below 0.05. The results are summarized in Table 1.

```{r}
#| warning: false
#| message: false
#| echo: false

# Extract coefficient
estimates <- summary(model_coded)$coefficients

# Extract beta estimates
beta_hat <- coef(model_coded)[c("x1", "x2", "x3", "x4", "x5", "x6", "x7")]

# Calculate main effect and interactions
effect_estimates <- 2 * beta_hat

# Estimated variance of effects
m <- 3
k <- 3
sigma2 <- summary(model_coded)$sigma ^ 2
var_effect <- (4 * sigma2) / (m * 2 ^ k)
se_effect <- sqrt(var_effect)

# Degrees of freedom for error
df <- summary(model_coded)$df[2]

# t-critical value for 95% CI
t_crit <- qt(0.975, df)

# Compute confidence intervals
lower <- effect_estimates - t_crit * se_effect
upper <- effect_estimates + t_crit * se_effect

# Final table
effects_table <- data.frame(
  Effect = c(
    "Temperature",
    "Sugar Type",
    "Stirring",
    "Temperature × Sugar",
    "Temperature × Stirring",
    "Temperature × Stirring",
    "Temperature × Sugar × Stirring"
  ),
  Estimate = round(effect_estimates, 2),
  Var = round(var_effect, 2),
  CI_Lower = round(lower, 2),
  CI_Upper = round(upper, 2),
  p_value = signif(estimates[c("x1", "x2", "x3", "x4", "x5", "x6", "x7"), "Pr(>|t|)"], 3)
)

# Display
knitr::kable(effects_table, caption = "Estimated main and interaction effects from the 2^3 factorial design, including variance, 95% confidence intervals, and p-values. Stirring and Temperature had the strongest effects on sugar dissolution time, while Sugar Type showed no statistically significant influence.")

```

The most influential factor was Stirring, with an estimated effect of $-804.27$ seconds. This indicates that manually stirring the solution drastically reduces the time required for sugar to dissolve. Temperature also had a substantial impact: with an effect of $-440.13$ seconds, hot water significantly accelerates dissolution compared to cold water. These large negative values confirm the intuitive physical expectations regarding agitation and thermal energy.

Sugar Type, on the other hand, was not statistically significant ($p = 0.783$). Its estimated effect was small and the 95% confidence interval included zero. This suggests that, in the context of this experiment, the difference in crystal structure and composition between white and brown sugar does not significantly affect dissolution time.

The interaction between Temperature and Stirring was statistically significant ($p < 0.05$). The estimated ineraction effect is 377.04, indicates that the effectiveness of stirring varies depending on water temperature. In particular, stirring has a much larger effect in cold water, where the natural dissolution process is slower. 

Other interaction terms, including Temperature × Sugar and Sugar × Stirring, were not significant. Their effects were small, and confidence intervals contained zero, and have p_value greater than 0.05, indicating no strong evidence of combined effects. The three-way interaction was also non-significant ($p = 0.374$), suggesting that the combined influence of all three factors does not go beyond the observed two-way interactions.


## Interaction Plots
To support these statistical findings, three interaction plots were generated to visualize the mean dissolution time for each two-factor combination.

```{r}
#| warning: false
#| message: false
#| echo: false
#| fig.width: 12
#| fig.height: 3

data_long <- data %>%
  mutate(
    Temp = factor(Temp, labels = c("Cold", "Hot")),
    Stirring = factor(Stirring, labels = c("No", "Yes")),
    Sugar = factor(Sugar, labels = c("White", "Brown"))
  )

# Set layout: 1 row, 3 plots side by side
par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))

# First: Temp × Sugar
interaction.plot(
  x.factor = data_long$Temp,
  trace.factor = data_long$Sugar,
  response = data_long$Time_to_Dissolve,
  xlab = "Temperature",
  ylab = "Mean Dissolution Time (s)",
  trace.label = "Sugar Type",
  main = "Temperature × Sugar",
  col = c("blue", "orange"),
  lwd = 2
)

# Second: Temp × Stirring
interaction.plot(
  x.factor = data_long$Temp,
  trace.factor = data_long$Stirring,
  response = data_long$Time_to_Dissolve,
  xlab = "Temperature",
  ylab = "",
  trace.label = "Stirring",
  main = "Temperature × Stirring",
  col = c("blue", "orange"),
  lwd = 2
)

# Third: Sugar × Stirring
interaction.plot(
  x.factor = data_long$Sugar,
  trace.factor = data_long$Stirring,
  response = data_long$Time_to_Dissolve,
  xlab = "Sugar Type",
  ylab = "",
  trace.label = "Stirring",
  main = "Sugar × Stirring",
  col = c("blue", "orange"),
  lwd = 2
)

```
\begin{center}
\textbf{Figure 1.} Interaction plots to visualize main effects and interactions. Plots indicating that stirring has a much larger effect under cold conditions, and there is no significant interaction between Sugar Type and Stirring, as well as between Temperature and Sugar Type.
\end{center}

The Temperature × Stirring plot shows clearly diverging lines, confirming that stirring is far more effective under cold conditions. The Sugar × Stirring plot displays near-parallel lines, consistent with the lack of statistical interaction. The Temperature × Sugar plot shows slight crossing, suggesting a possible interaction trend, but this is not supported by the statistical analysis.

# Conclusions
This experiment aimed to investigate how water temperature, sugar type, and stirring affect the dissolution time of granulated sugar using a replicated $2^3$ factorial design. The results demonstrate that both water temperature and stirring have substantial and statistically significant effects. Stirring reduced dissolution time by over 800 seconds on average, making it the most influential factor. Temperature followed closely, with hot water reducing dissolution time by more than 400 seconds. 

The interaction between temperature and stirring was also statistically significant. Stirring had a stronger effect under cold conditions, suggesting that agitation can partially compensate for the reduced kinetic energy in colder water. On the other hand, sugar type did not show any significant effects, meaning that the difference between white and brown sugar did not influence the dissolution time in the conditions tested.

Overall, the factorial design enabled efficient estimation of individual and joint effects while controlling for variability. However, several limitations were identified in the course of conducting this experiment. First, during trials without stirring, especially in the hot water condition, the water temperature may have decreased over time despite being initially heated and covered, potentially reducing consistency across trials. Second, while effort was made to standardize how sugar was poured into the container, particularly for brown sugar, clumping and uneven dispersion may have introduced variability. Third, although stirring was performed manually at an estimated frequency, it was difficult to maintain perfect consistency across trials. Lastly, although transparent containers were used, it was still somewhat subjective to determine the exact moment when sugar had fully dissolved, particularly for fine residual particles.

Future experiments could benefit from more controlled stirring mechanisms (e.g., magnetic stirrers), pre-calibrated temperature tracking to monitor water cooling over time, and automated visual sensors to detect complete dissolution. Additionally, increasing the number of trials and improving consistency across factors could lead to even more reliable results.